How to \emph{not} write Virtual Machines for Dynamic Languages

### 2007-07-12 talk/dyla2007/dyla.tex
Start to answer to the reviewer's comments.

The crux of the argument is that VMs for dynamic languages should not be
written by hand!  The PyPy project is the justification,
proving that the approach is
feasible in practice.  Just as importantly, it also brings new insights
and concrete benefits in term of flexibility and performance that go
beyond the state of the art.

@@

\begin{itemize}
\item
\emph{Better performance:} So far it seems that performance of highly dynamic
languages is not actually significantly improved on OO VMs. 
Jython is around 5
times slower than CPython, for IronPython (which
gives up on at least one feature -- frame objects -- to improve performance)
the figures vary but it is mostly
within the same order of magnitude as CPython. The most important reason for
this is that the VM's JIT compilers are optimized for specific usage patterns
that are common in the primary language of the OO VM. To achieve good speeds, the
language implementers would have to carefully produce code that matches these
usage patterns, which is not a simple task.

\item
\emph{Better GCs:} While this is obvious in theory, OO VMs tend to have a much
higher memory overhead to start with (XXX ref)

>>>>

The crux of the argument is that VMs for dynamic languages should not be
written by hand!  The PyPy project \cite{pypy} is the justification,
proving that the approach is
feasible in practice.  Just as importantly, it also brings new insights
and concrete benefits in term of flexibility and performance that go
beyond the state of the art.

PyPy contains a Python interpreter implemented in Python, from which
Python VMs can be generated.  The reader is referred to
\cite{pypyvmconstruction} for a technical presentation.  Let us emphasis
again that the argument we make in the present paper is \emph{not} that
VMs for dynamic languages should be written in their own host language
(as many projects like Squeak \cite{Squeak} and others have done) but
instead that VMs should not be \emph{written} in the first place -- they
should be generated from simple interpreters written in any suitable
high-level language.

@@

\begin{itemize}
\item
\emph{Better performance:} So far it seems that performance of highly dynamic
languages is not actually significantly improved on OO VMs. 
Jython is around 5
times slower than CPython; for IronPython (which
gives up on at least one feature -- frame objects -- to improve performance)
the figures vary but it is mostly
within the same order of magnitude as CPython. The most important reason for
this is that the VM's JIT compilers are optimized for specific usage patterns
that are common in the primary language of the OO VM. To achieve good speeds, the
language implementers would have to carefully produce code that matches these
usage patterns, which is not a simple task.

\item
\emph{Better GCs:} While this is obvious in theory, OO VMs tend to have a
higher memory overhead to start with.  For example, an instance of Sun's
Java VM which just loaded Jython consumes XXX MB of non-shared memory
(XXX for the JVM and an additional XXX after loading Jython), while a
CPython process fits in XXX MB.
